---
title: "Soc 722"
execute: 
  echo: true
  warning: false
format: 
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
editor: source
revealjs-plugins:
  - revealjs-text-resizer
---

# Logistics

## Schedule

We will normally meet Tuesdays. Please note that we will be meeting on the following **Thursdays** (not Tuesdays) because of my travel schedule:

- today
- January 30
- February 6
- February 13
- March 20

## Final exam

The exam will be **remote** on April 30 from 9am-12pm.

##

:::{.r-fit-text}
Questions?
:::

# Introduction

## Objective

This is the first in a two-course sequence designed to help you become competent quantitative researchers in sociology.

This includes learning proper **decision making**, **explanation**, **computation**, **visualization**, and **interpretation**.

# Data and Variables

## Data structure

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(gapminder)
library(tinytable)
library(knitr)
library(kableExtra)

theme_set(theme_light())
```

```{r}
#| echo: false
data(gapminder)
head(gapminder, n = 10) |> 
  kbl("html") |>
  kable_styling(font_size = 24)
```

Tidy format: columns contain **variables**, each row is an **observation**.

## Untidy data

```{r}
#| echo: false

untidy <- gapminder |> 
  pivot_wider(values_from = c(lifeExp, pop, gdpPercap),
              names_from = year)

head(untidy, n = 12) |> 
  kbl("html") |>
  kable_styling(font_size = 24)
```


## Types of variables

|              |                                    |
|--------------|------------------------------------|
| **Ratio**    | dollars; points (e.g., basketball) |
| **Interval** | degrees Celsius                    |
| **Ordinal**  | clothing sizes; Likert scales      |
| **Nominal**  | race; sex; country                 |

: {.striped}

The first two types are **continuous** or **numeric**. The second two types are **categorical**. Ordinal variables are often treated as numeric and this is usually fine.

## 

Let's investigate this using the `gapminder` data. First of all, we'll keep only the most recent (2007) data.

```{r}
d <- gapminder |>               
  filter(year == max(year)) |> # keep 2007
  select(-year)                # don't need column
```

## 

```{r}
#| echo: false

head(d, n = 10) |> 
  kbl("html") |>
  kable_styling(font_size = 24)
```

What kinds of variables are these?

## The origins of "statistics"

The word _statistics_ comes from the fact that it was information about _the state_. We'll focus on information like this for now rather than thinking about samples of individuals.

## Visualization basics

Consider two types of plots

- univariate plots

- bivariate plots

These are also types of **distributions.**

## Univariate plots

## 

```{r}
# density plot
ggplot(d,
       aes(x = gdpPercap)) +
  geom_density()
```

## 

```{r}
# histogram
ggplot(d,
       aes(x = gdpPercap)) +
  geom_histogram(binwidth = 5000,
                 boundary = 0,
                 color = "white")
```

## 

```{r}
# histogram
ggplot(d,
       aes(x = lifeExp)) +
  geom_histogram(binwidth = 5,
                 boundary = 0,
                 color = "white")
```

##

```{r}
# bar graph
ggplot(d,
       aes(x = continent)) +
  geom_bar()
```

## Bivariate plots

## 

```{r}
# scatterplot
ggplot(d,
       aes(x = gdpPercap,
           y = lifeExp)) +
         geom_point()
```

## 

```{r}
# bar graph (bivariate)
d |> 
  group_by(continent) |> 
  summarize(GDP = mean(gdpPercap)) |>
  ggplot(aes(x = continent,
             y = GDP)) +
  geom_bar(stat = "identity")
```

## Time plots (bivariate)

Let's go back to the full data and look at trends in Oceania.

```{r}
oceania_plot <- gapminder |> 
  filter(continent == "Oceania") |> 
  ggplot(aes(x = year,
             y = lifeExp,
             group = country,
             color = country)) +
  geom_line()
```

##

```{r}
oceania_plot
```

##

:::{.r-fit-text}
Your turn!
:::
